Comparison of MCMC algorithms in Stochastic Volatility Models

### 2023-09-24 manuscript/main.tex
Updated examples and added research context

\begin{abstract}.  
    The objective of this research is to compare the computational methods used to estimate Bayesian stochastic volatility models. A simulation study will compare the estimation strategies detailed in the stochastic volatility literature with Hamiltonian Monte Carlo and their ability to sample from the model's posterior distribution. Specifically, Simulation Based Calibration (SBC) is used to check whether these sampling strategies are returning efficient and well calibrated posterior estimates. Key metrics of interest are the effective sample size to check the efficiency of the algorithm and tests of uniformity to assess the calibration of the posteriors. This will determine which algorithm is better at estimating stochastic volatility models based on the efficiency and accuracy of the sampling strategy.
\end{abstract}

@@

\section{Introduction}

- Should make reference to workflow, and reparameterisation of SV.

\section{Background and Literature Review}

    \subsection{Stochastic Volatility}
        Stochastic volatility models are used in financial econometrics to model the variance or volatility of returns in a financial instrument. The return of a financial asset often exhibits patterns in the variance which can be modelled to improve prediction and estimates of risk or tail events. Unlike other volatility models such as ARCH or GARCH, stochastic volatility models treat the variance as a random variable. Furthermore, these models have a state space representation since the variance is expressed as a latent stochastic process. State space models can be difficult to estimate since there is a latent state for every data observation in the time series, which means there are often more parameters than data points. These models can be estimated using frequent frameworks but this research will focus on Bayesian estimation strategies.

        \citet{kim1998stochastic} estimate a discrete time, univariate stochastic volatility model using Bayesian methods which is outlined below. $y_t$ is the mean corrected returns of some asset for equally spaced intervals t. $\beta$ is a constant scaling factor which is also defined as $exp(\mu / 2)$ representing instantaneous volatility. $h_t$ is log volatility, where $h_1$ is a draw from a stationary distribution and the state equation $h_{t+1}$ follows a stationary process governed by the autoregressive parameter $\phi$ such that $|\phi|<1$. This autoregressive parameter represents the persistence or "stickiness" of log volatility and the dispersion parameter $\sigma_{\eta}$ is the constant variance of the states. $\epsilon_t$ and $\eta_t$ are standard normal white noise shocks and are uncorrelated with each other.

@@

    \subsection{Gaussian Mixture Approximation}
        Kim, Shephard and Chib (KSC) sample the posteriors of the stochastic volatility model using a mix of conjugate posterior distributions, Metropolis Hastings and the Kalman Filter and smoother\footnote{In this research the exact software to apply the simulation smoother is unavailable, so a more recent simulation smoother is used which is based on the software written by the same author.} \citep{dejong1995}.

    \subsection{Hamiltonian Monte Carlo}
        Since the paper was written, new MCMC algorithms have been developed and enabled the estimation of richer and more complicated models. Specifically, Hamiltonian Monte Carlo is a MCMC algorithm which has become widely available for efficiently sampling from sophisticated models. Hamiltonian Monte Carlo, originally called Hybrid Monte Carlo, was developed in the physics literature \citep{duane1987hybrid} before being applied in the statistics literature by Radford Neal through his works in Bayesian Neural Networks \citep{neal1995bayesian} and statistical computing \citep{neal2011mcmc}. The algorithm has since become widely available through open source development projects such as Stan \citep{stan} and PyMC \citep{pymc2023}.

@@

    \subsection{Simulation Design}
        Simulation Based Calibration (SBC) is a technique that checks the calibration of posterior estimates generated by MCMC algorithms. SBC is conducted by comparing the distribution of rank statistics to the discrete uniform distribution which arises when an algorithm is correctly calibrated. The procedure starts by taking draws from the prior distribution and creating datasets implied by each draw. Rank statistics are then calculated on the posterior samples conditional on the simulated data.

@@

    \subsection{Metrics}
        The key metrics and diagnostics to compare the performance of these methods are the effective sample size (ESS), rank statistics, and chi-squared test statistics.

        ESS measures the efficiency of the MCMC sampler. It calculates the number of effectively independent draws from the posterior draws generated by a Markov chain. A poor ESS can arise from high autocorrelation in the Markov chain which leads to highly dependent samples. An efficient MCMC algorithm takes less resources (for example, time and number of draws) to get a representative sample of the target distribution. If a MCMC algorithm possesses higher ESS for the majority of its parameters (relative to another strategy), then we may conclude that this method is a more efficient sampler (conditional on the model).

        Rank statistics as described in the SBC section are used to evaluate the calibration of a posterior. Histograms will be used to evaluate the distribution of rank statistics. If a posterior is well calibrated then it is expected that the histogram is uniform.

@@

        Chi squared statistics can be used to test for uniformity and thus calibration of all the parameters in the model.

@@

        The parameterisation of a model can affect the performance of a MCMC algorithm when sampling from models with complex posterior geometries. An example of this is Neal's funnel \citep{neal2003slice} where the Hamiltonian Monte Carlo sampler encounters performance issues in hierarchical models and produces biased samples.

        The stochastic volatility model as described in section 2.1 is follows a "centered" parameterisation. This describes the central location of the latent state parameters which are centered around the mean and lag of the log volatility $\mu +\phi(h_t - \mu)$. A model can be reparameterised such that the states are sampled on a distribution centered on 0 (i.e "non centered") which are later transformed to have the correct mean.  It is useful to compare model parameterisations since sampling strategies may be sensitive to the posterior geometry.

        \subsubsection{Non centered Hamiltonian Monte Carlo}
        First sample from a standard normal distribuition and multiply by the variance of the log volatility.

@@

        The states are sampled with the mean centered at zero. Then apply the appropriate rescaling to get samples from the log volatility.

@@

        \subsubsection{Non centered Gaussian Mixture}
        The non centered model for the Gaussian approximation is expressed slightly differently due to the use of the Kalman Filter. While HMC can sample from the joint posterior directly with state vectors centered at zero, applying the Kalman Filter requires the state equation to be a function of the state variable.

@@

        Expand the state equation:

        $$
        \begin{aligned}
        h_{t+1} =& \space \mu +\phi(h_t - \mu) + \sigma_{\eta} \eta_t  \\
        =& \space \mu + \phi h_t - \phi\mu + \sigma_{\eta} \eta_t  \\
        =& \space \mu (1- \phi) + \phi h_t + \sigma_{\eta} \eta_t  \\
        \end{aligned}
        $$

        Take out the intercept and put it in the measurement equation:

        $$
        \begin{aligned}
        y^{\ast}_t =& h_t + \mu (1- \phi) + z_t \\
        h^{\ast}_{t+1} =& \space \phi h_t + \sigma_{\eta} \eta_t  \\
        \end{aligned}
        $$

@@

        \subsubsection{Prior Predictive check}
        It is worth noting that the centered and non-centered parameteristions are the same stochastic volatility model. Different parameteristions express the same mathematical models in ways that may make it easier or harder for a given algorithm to sample.

@@

        % \begin{figure}
        %     \centering
        %     \includegraphics[scale=0.1]{ppc_h2.png}
        %     \caption{Prior predictive samples of second state variable}
        % \end{figure}

>>>>

\begin{abstract}
    This research compares the computational methods used to estimate Bayesian stochastic volatility models. The estimation strategies outlined in the stochastic volatility literature and Hamiltoninan Monte Carlo are assesssed in their ability to sample from the model's posterior distribution. Specifically, Simulation Based Calibration (SBC) is used to check whether these MCMC algorithm are returning efficient and well calibrated posterior estimates. Key metrics of interest are the effective sample size to check the efficiency of the algorithm and tests of uniformity to assess the calibration of the posteriors. This will determine which method is better at estimating stochastic volatility models based on the efficiency and accuracy of the sampling strategy. Results reveal that Hamiltonian Monte Carlo provides more efficient and calibrated posterior estimates conditional on the non centered paramterisation of the stochastic volatility model. 
\end{abstract}

@@

\section{Introduction}
    The stochastic volatility model is used in financial econometrics to model the behaviour of financial instruments. It explicitly treats the variance as a latent random variable. These are typically expressed as non linear Gaussian state space models and are difficult to estimate using classical methods. The likelihood is unavailable in closed form and there are at least as many variables as data points. Bayesian approaches to estimating these models rely on  Markov Chain Monte Carlo (MCMC) techniques to sample from the target joint posterior of these high dimensional latent parameter spaces.

    \citet{kim1998stochastic} propose a method for sampling this model using a combination of Kalman Filters, Gibbs sampler, and Metropolis Hastings over 20 years ago.  Since then, there have been many advances in statistical computing to estimate complex, high dimensional Bayesian models. In particular, the recent availability of Hamiltonian Monte Carlo, a variant of the Metropolis Hastings algorithm, is an advancement of MCMC for efficient sampling of complex probabilistic models. The adoption of such new techniques have been made widely available through the development of various open source libraries such as the Stan programming language \citep{stan} and the PyMC library \citep{pymc2023}. Other algorithms have also improved the speed of estimating high dimensional models using approximations of the posterior such as integrated nested Laplace approximation \citep{rue2009approximate}.

    These conceptually and practically different techniques to estimate complex models did not exist 20 years ago (or perhaps more accurately, were not easily accessible or implemented 20 years ago). As the development of new algorithms rapidly increase, so does the need to develop new methods to assess their output. Developments in our statistical workfow are required to test new algorithms as well as compare computational strategies used to estimate increasingly complex models.

    This research conducts a simulation study to assess the algorithms used to estimate stochastic volatility models. The first algorithm replicates Kim Sherphard and Chib's (KSC) Gaussian mixture approximation which is applied on a transformed stochastic volatility model with a log chi squared error distribution. Their sampling method uses the aforementioned Kalman Filter estimating the latent states, conjugate posterior distributions and the Metropolis Hastings algorithm. The second algorithm is the Hamiltonian Monte Carlo (HMC) algorithm as implemented in the Stan programming language. The objective is to analyse the efficiency of the sampling approach and their ability to return calibrated posterior estimates.

    Comparisons of these algorithms are conducted by Simulation Based Calibration \citep{talts2018validating}. Parameters are drawn from the prior distribution and used to create datasets from the generative (stochastic volatility) model which are sampled using a MCMC method. Repeating this process multiple times gives insight to how well the sampling approach can estimate the true parameters governing the data generating process on average. The key diagnostic metrics are the effective sample size to measure the efficiency of the algorithm and visual tests of uniform rank statistics to assess calibration of the posterior estimates.

    Additionally, two different parametiersations of the stochastic volatility model are considered for each sampling method. These are described as either 'centred' or 'non-centred' in location. An algorithm may be sensitive to different parameterisations of the same model which may impact MCMC efficiency as described in \citet{strickland2008parameterisation}. The key contribution of this research is to use simulations to assess whether a MCMC algorithm can return calibrated posterior estimates as well as the efficiency under the different parametiersations.

    This paper is structured as follows. Section 2 provides the context around this research, namely the stochastic volatility model examined and the limitations of single simulation studies and MCMC convergence diagnostics. Section 3 describes the simulation design and diagnostic metrics. Section 4 details the sampling approaches. Section 5 discusses the results and limitations. Finally, section 6 concludes and provides points for further reserach.

\section{Resarch Context}

\subsection{Stochastic Volatility}
    The model of interest is the discrete time, univariate stochastic volatility model estimated by \citet{kim1998stochastic} using Bayesian methods. $y_t$ is the mean corrected returns of some asset for equally spaced intervals t. $\beta$ is a constant scaling factor which is also defined as $exp(\mu / 2)$ representing instantaneous volatility. $h_t$ is log volatility, where $h_1$ is a draw from a stationary distribution and the state equation $h_{t+1}$ follows a stationary process governed by the autoregressive parameter $\phi$ such that $|\phi|<1$. This autoregressive parameter represents the persistence or "stickiness" of log volatility and the dispersion parameter $\sigma_{\eta}$ is the constant variance of the states. $\epsilon_t$ and $\eta_t$ are standard normal white noise shocks and are uncorrelated with each other.

@@

\subsection{Diagnostic limitations on real data}
    The stochastic volatility model can be estimated using a variety of sampling strategies and MCMC algorithms. Convergence metrics are often used to check the performance of the MCMC chains, such as the effective sample size and the $\hat{R}$. These diagnostics provide evidence on whether an MCMC has failed to converge onto the target posterior distribution.

    Convergence diagnostics are useful for identifying when Bayesian computation fail on real data. However, confounding issues may arise when attempting to diagnose the cause of computational problems. Real data is generated from an unknown data generating process. That is, the true parameter or model is unobservable. Therefore, failed diagnostic checks could arise from either inocrrect model specification or issues in the sampling algorithm or both.

    Furthermore, different sampling strategies and MCMC tools could provide different posterior estimates for the same model. Assuming no issues in computation, these diagnostic measures do not provide any evidence to which estimate is closer to the truth.

    To illustrate this problem, the stochastic volatility model is estimated with two different MCMC algorithms on real data. The data are the de-meaned log returns of the 2023 January-September S\&P 500 Index. Below are the marginal posterior estimates for the static parameters $\mu$, $\phi$ and $\sigma^2$ and corresponding summary statistics.

    \begin{figure}[h]
        \centering
        \includegraphics[scale=0.1]{motivating_example/real_data_ex.png}
        \caption{Posterior samples from two MCMC algorithms using S\&P 500 data}
    \end{figure}

    These estimates show the marginal posteriors of each parameter sharing similar shapes. $\phi$  has a fatter left tail for and $\sigma^2$ has heavier right tails for sampling method 1 (SM1). Furthermore, SM2 has higher peaks at their mode relative to SM1 with a tighter spread. Based on this information alone, there is no certainty in whether SM1 or SM2 is has a more correct estimate conditional on data and model. There are other methods for model or algorithmic selection in this context (for example, out of sample predictive performance), but for parameter estimation it is unclear which one should be selected.

\subsection{Diagnostic limitations on a single simulation}
    Diagnosing computational problems on real data is difficult due to confounding factors. A strategy around this is to evaluate a model and algorithm on simulated data. One approach is to simulate data from the proposed model using known parameters. Then, fit the model on the simulated data and see if the model can recover the true parameters. This gives us the benefit of defining the true parameters of the data generating process to be estimated. If the model and algorithm cannot adequately capture the true parameter, then we cannot be confident that it will provide reliable estimates on real data.

    Furthermore, as discussed in \citet{gelman2020bayesian}, fitting models on simulated data is the only way we can check inference on latent variables. This is critical in Stochastic Volatility since the underlying framework is a state space model with latent log volatility parameters. Latent variables are unobserved in real data and are only estimated in the context of the model. Simulation gives control over the data generating process which reveals what the model can infer about the latent variables.

    Simulations enable us to check whether our models and algorithms can estimate the true data generating process, however, there are limitations to what can be learned from a single simulation. There is always a small probability that the true parameter is in the tails of any estimated posterior distribution. Or put differently, there is a 1\% chance the true parameter or any random draw exists outside a 99\% credible interval. \cite{talts2018validating} make the point that a single simulation does not provide sufficient information about the inference made by an algorithm. As discussed in their paper, a single simulation may conclude "that an incorrectly coded analysis worked as desired, while a correctly coded analysis failed".

    The below results are from a single simulation where data is generated and sampled from the stochastic volatility model with known parameters.

    \begin{figure}[h]
        \centering
        \includegraphics[scale=0.1]{motivating_example/single_sim.png}
        \caption{Posterior samples from known data genearting process. Vertical solid line represents true parameter and dotted lines are 95\% credible intervals around the mean}
    \end{figure}

    The 95\% credible intervals for $\sigma$ and $\phi$ cover the true parameter. The true parameter for $\mu$ however, is deep in the tails and outside the interval. Such an analysis may incorrectly conclude that the model fails to adequatly estimate the $\mu$ parameter. However, it may be the case that the posterior distribution is correctly calculated using this algorithm and the results may be due to the features of this specific simulated dataset.

\subsection{Research Goal}
    The objective of this research is to design a simulation study to evaluate the calibration of algorithms used to sample Stochastic Volatility models. Indeed, there are limitations to evaluating MCMC algorithms based on fits to real data and single simulations. To check the calibration of an algorithm, simulations over multiple "true parameters" are required. This methodology is discussed in the next section.

    Additionally, two sampling strategies will be assessed using the proposed simulation design as well as the performance of the samplers under different parameterisations of the stochastic volatility model. Not only will the calibration of any one algorithm be assessed, but also compared across sampling strategies to determine which MCMC approach is most suitable for estimating this model.

@@

    \subsection{Simulation Design}
        Simulation Based Calibration (SBC) checks the calibration of posterior estimates generated by MCMC algorithms. SBC is conducted by comparing the distribution of rank statistics to the discrete uniform distribution which arises when an algorithm is correctly calibrated. The procedure starts by taking draws from the prior distribution and creating datasets implied by each draw. Rank statistics are then calculated on the posterior samples conditional on the simulated data.

@@

    \subsection{Evaluation Metrics}
        The key metrics and diagnostics to compare the performance of these methods are the effective sample size (ESS), rank statistics, and chi-squared test statistics.

        \subsubsection{Effective sample size}
            ESS measures the efficiency of the MCMC sampler. It calculates the number of effectively independent draws from the posterior draws generated by a Markov chain. A poor ESS can arise from high autocorrelation in the Markov chain which leads to highly dependent samples. An efficient MCMC algorithm takes less resources (for example, time and number of draws) to get a representative sample of the target distribution. If a MCMC algorithm possesses higher ESS for the majority of its parameters (relative to another strategy), then there is evidence that this method is a more efficient sampler.

        \subsubsection{Rank statistics and chi squared statistics}
            Rank statistics as described in the SBC section are used to evaluate the calibration of a posterior. Histograms will be used to evaluate the distribution of rank statistics. If a posterior is well calibrated then it is expected that the histogram is uniform.

@@

            A perfectly discrete uniform distribution will return a $\chi^2$ statistic of zero. That is, the number of rank statistics in each bin is equal to the expected number conditional on the number of bins and observations. Instead of performing multiple hypothesis tests for each parameter, the distribution of $\chi^2$ statistics is compared across all parameters and simulations. This will give a high level summary of calibration and overall performance of the algorithms.

\section{MCMC Strategies}

    \subsection{Sampling method 1}
        (KSC) sample the posteriors of the stochastic volatility model using a mix of conjugate posterior distributions, Metropolis Hastings and the Kalman Filter and smoother\footnote{In this research the exact software to apply the simulation smoother is unavailable, so a more recent simulation smoother is used which is based on the software written by the same author.} \citep{dejong1995}.

    \subsection{Sampling method 2}
    % Need much more detail here about how it works
        Since the paper was written, new MCMC algorithms have been developed and enabled the estimation of richer and more complicated models. Specifically, Hamiltonian Monte Carlo is a MCMC algorithm which has become widely available for efficiently sampling from sophisticated models. Hamiltonian Monte Carlo, originally called Hybrid Monte Carlo, was developed in the physics literature \citep{duane1987hybrid} before being applied in the statistics literature by Radford Neal through his works in Bayesian Neural Networks \citep{neal1995bayesian} and statistical computing \citep{neal2011mcmc}. The algorithm has since become widely available through open source development projects such as Stan \citep{stan} and PyMC \citep{pymc2023}.

@@

        The parameterisation of a model can affect the performance of a MCMC algorithm when sampling from models with complex posterior geometries. An example of this is Neal's funnel \citep{neal2003slice} where the Hamiltonian Monte Carlo sampler encounters performance issues in hierarchical models and produces biased samples. Efficiency improvements of MCMC algorithms for state space models under different parameterisations are also explored \citet{strickland2008parameterisation}.

        The stochastic volatility model as described in section 2.1 is follows a "centered" parameterisation. This describes the central location of the latent state parameters which are centered around the mean and lag of the log volatility $\mu +\phi(h_t - \mu)$. A model can be reparameterised such that the states are sampled on a distribution centered on 0 (i.e "non centered") which are later transformed to have the correct mean. The following re-parameterisations are explored as part of the simulation study.

        \subsubsection{Non centered Hamiltonian Monte Carlo}
        First sample from a standard normal distribuition and multiply by the variance of the log volatility. The state variable is sampled with mean centered on zero and variance equal to $\sigma_{\eta}^2$

@@

        Then apply the appropriate rescaling to get samples from log volatility.

@@

        \subsubsection{Non centered Gaussian Mixture}
        The non centered model for the Gaussian approximation is expressed slightly differently due to the use of the Kalman Filter. Non centered HMC samples from the joint posterior directly with state vectors centered at zero and transformed to return the correct log volatility estimates. Applying the Kalman Filter requires the state equation to be a function of the state variable with the average log volatility parameter $\mu$ entering the measurement equation.

@@

        Let $g$ be the demeaned state variable. Rewrite the demeaned state equation $g_{t+1}$ to be non centered in location by subtracting average volatility $\mu$.

        $$
        \begin{aligned}
        g_t =& h_t - \mu \\
        g_{t+1} =& \phi g_t + \sigma_{\eta}\eta_{t}
        \end{aligned}
        $$

        Return average volatility into measurement equation and rewrite as a function of the non centered state equation:

        $$
        \begin{aligned}
        y^{\ast}_t =& g_t + \mu + z_t \\
        g_{t+1} =& \phi g_t + \sigma_{\eta}\eta_{t}
        \end{aligned}
        $$

@@

    \subsubsection{Prior Predictive check}
        It is worth noting that the centered and non-centered parameteristions are the same stochastic volatility model. Different parameteristions express the same mathematical models in ways that make it easier or harder for a given algorithm to sample.

@@

        \begin{figure}[h]
            \centering
            \includegraphics[scale=0.1]{ppc_h2.png}
            \caption{Prior predictive samples of second state variable}
        \end{figure}
